---
title: "Creating bias-corrected time series of sea temperature"
author: "Kelly Ortega-Cisneros, Denisse Fierro-Arcos"
date: "2024-10-24"
output: html_document
---
In this notebook, we describe the process to create time series of bias corrected temperature following the guidelines from Ortega-Cisneros et al., (2024).

We will use data data downloaded from our FishMIP Input Explorer app, specifically, the data from the 'Model outputs against observations' tab for three case study regions. The data is downloaded as a folder containing four files. You will need to unzip the folder before running this script to access the .parquet files.

#Loading relevant libraries

```{r setup, include=FALSE}
.libPaths("/home/user/R/x86_64-pc-linux-gnu-library/4.3")
library(nanoparquet)
library(dplyr)
library(Rarr)
library(tidyr)
library(terra)
library(sf)
library(ggplot2)
library(lubridate)
library(cowplot)
```

### Loading parquet files for the southern Benguela downloaded from the FishMIP Shiny app
```{r }
sb_g <- file.path(
  data,
  "gfdl-mom6-cobalt2_obsclim_thetao_15arcmin_southern-benguela_mthly_clim_mean_1981_2010.parquet") |> 
  read_parquet()
sb_woa <- file.path(
  data,
  "regridded_woa_southern-benguela_month_clim_mean_temp_1981-2010.parquet") |> 
  read_parquet()
sb_areacello <- file.path(
  data,
  "gfdl-mom6-cobalt2_areacello_15arcmin_southern-benguela_fixed.parquet") |> 
  read_parquet()
sb_depth <- file.path(
  data,
  "gfdl-mom6-cobalt2_obsclim_deptho_15arcmin_southern-benguela_fixed.parquet") |> 
  read_parquet() 
```

### Exploring file contents
```{r }
parquet_column_types(sb_g)
parquet_column_types(sb_woa)
parquet_column_types(sb_areacello)
parquet_column_types(sb_depth)
```

### Processing temperature data to calculate climatologies
```{r }
#Plotting temperature data for the upper most layer from GFDL
sb_g |>
  filter(depth == min(depth)) |>
  ggplot(aes(lon, lat, fill = vals)) +
  geom_tile()

#Calculating height of depth bins - This step is needed because the depth value
#includes the depth of the water column up to that grid cell. If weighting by
#depth, we should only include the depth for the grid cell (i.e., height of the
#grid cell).
depth_bins <- sb_g |> 
  filter(depth < 501) |> 
  distinct(depth) |> 
  #Calculating the difference between each depth bin to get the height of the
  #grid cell
  mutate(depth_height = depth-lag(depth, default = 0))

# Calculating GFDL climatologies from 1981-2010
sb_g_clims <- sb_g |>
  #Keeping relevant columns only
  dplyr::select(lat:vals) |> 
  rename(temperature = vals) |>
  filter(depth < 501) |>
  drop_na(temperature) |>
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |> 
  rename(area = vals) |> 
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  #Keeping relevant columns only
  dplyr::select(!standard_name:units)
  
  
#Reading shapefile to perform data extraction for the model polygons
cb <- st_read("/data/model_regions_v3_geo3.shp")

#Create raster from data frame
ras <-  sb_g_clims |>
  ungroup() |>  
  #A single date and layer is needed as a sample
  filter(layer == min(layer) & month == min(month)) |> 
  #Only these three columns are needed to create a raster
  dplyr::select(lon, lat, area) |> 
  #Create raster
  rast(type = "xyz", crs = "epsg:4326")

#Create raster mask
cb_mask <- rasterize(cb, ras, field = "BOX_ID", background = NA) |> 
  #Transform to data frame
  as.data.frame(xy = T) |> 
  rename(lon = x, lat = y)
#This contains a data frame identifying the group that each cell belongs to

#Join data frames
sb_g_clims <- sb_g_clims |>
  full_join(cb_mask, by = c("lat", "lon"))

#Finalise the data extraction for the model polygons
sb_g_clim_box <- sb_g_clims |>
  group_by(layer, month, BOX_ID)|>
  #Using volume as weights in mean calculation
  drop_na(temperature) |>
  summarise(w_temp_gfdl = weighted.mean(temperature, volume, na.rm = TRUE)) |> 
  #Transforming month column to factor
  mutate(month = factor(month, levels = month.name)) |>
  arrange(as.numeric(BOX_ID),month)

#Plotting temperature data for the upper most layer from WOA climatology
sb_woa |>
  filter(depth == min(depth)) |>
  ggplot(aes(lon, lat, fill = vals)) +
  geom_tile()

# Calculating WOA climatologies from 1981-2010 for the 0-50 m depth layer or realm of this model
sb_w_clim_box <- sb_woa |>
  #Keeping relevant columns only
  dplyr::select(lat:vals) |> 
  rename(temperature = vals) |>
  filter(depth < 501) |>
  drop_na(temperature) |>
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |>
  rename(area = vals) |> 
  # Joining cb_mask (data frame identifying the box that each cell belongs to) and the WOA dataset
  full_join(cb_mask, by = c("lat", "lon")) |>
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  dplyr::select(!standard_name:units) |> 
  #Excludes grid cells that are not present in GFDL
  drop_na(volume, temperature) |> 
  group_by(layer, month, BOX_ID) |>
  #Using area as weights in mean calculation
  summarise(w_temp_woa = weighted.mean(temperature, volume, na.rm = TRUE)) |>
  #Transforming month column to factor
  mutate(month = factor(month, levels = month.name)) |>
  arrange(BOX_ID, month)
```

### Loading GFDL obsclim Zarr file downloaded from the FishMIP Input Explorer app for the southern Benguela 
```{r}
sb_temp <- file.path(
  data,
  "gfdl-mom6-cobalt2_obsclim_thetao_15arcmin_southern-benguela_monthly_1961_2010.zarr")
```

### Exploring zarr file contents
```{r }
zarr_overview(sb_temp)

  # Path: thetao
  # Shape: 600 x 35 x 36 x 55
  # Chunk Shape: 600 x 5 x 36 x 55
  # No. of Chunks: 7 (1 x 7 x 1 x 1)
  # Data Type: float32
  # Endianness: little
  # Compressor: blosc
```

### Loading data

We will add the variable name to the end of the `zarr` path we defined above to load all data.
```{r }
temp_data <- read_zarr_array(file.path(sb_temp, "thetao"))
depth <- read_zarr_array(file.path(sb_temp, "depth_bin_m"))
lat <- read_zarr_array(file.path(sb_temp, "lat"))
lon <- read_zarr_array(file.path(sb_temp, "lon"))
time <- read_zarr_array(file.path(sb_temp, "time"))
```

### Checking dimensions of temperature data

```{r }
dim(temp_data)
#File dimensions are 600 (time) 35 (depth) 36 (lat)  55 (lon)

#plot slice 1 for time and depth
image(temp_data[1,1,,])
```

### Creating a data frame from all components

Important to check dimensions of dataset. Note that dimensions need to match the order of dimensions in the temperature data
```{r }
temp_sb <- cbind(expand.grid(time, depth, lat, lon), val = as.vector(temp_data))

#Rename the columns to reflect their contents
names(temp_sb) <- c("time", "depth", "lat", "lon", "temperature")
```

### Calculate monthly temperature time series from GFDL obsclim

```{r}
#Process data from GFDL obsclim 
sb_gfdl_box <- temp_sb |> 
  filter(depth < 501) |> 
  drop_na(temperature) |>
  mutate(date = as.Date("1961-01-01") %m+% months(round(time/30.417, 0))) |>
  #grouping depth bins into the four depth layers of the southern Benguela Atlantis
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |>
  rename(area = vals) |> 
  # Joining cb_mask (data frame identifying the box that each cell belongs to) and    the GFDL time series
  full_join(cb_mask, by = c("lat", "lon")) |>
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  dplyr::select(!standard_name:units) |> 
  #Excludes grid cells that are not present in GFDL
  drop_na(volume, temperature) |> 
  group_by(layer, date, BOX_ID) |>
  #Using area as weights in mean calculation
  summarise(w_temp_ts = weighted.mean(temperature, volume, na.rm = TRUE)) |>
  arrange(BOX_ID, date)
 
 
#plot the temperature time series ranging from January 1961 to December 2010
sb_gfdl_box |>
  filter(BOX_ID == 0 & layer == 1)|>
  ggplot(aes(date, w_temp_ts)) +
  geom_line() +
  labs(y = "Sea temperature 0-50 m depth for box 1 and layer 1") +
  theme_bw()
```

### Calculate bias corrected temperature time series 

```{r}
#Join monthly means with monthly climatologies from GFDL and WOA
sb_temp_corr_box <- sb_gfdl_box |> 
  mutate(month = factor(month.name[month(date)])) |> 
  left_join(sb_g_clim_box, by = c("layer", "month", "BOX_ID")) |> 
  left_join(sb_w_clim_box, by = c("layer", "month", "BOX_ID")) |> 
  #Apply correction
  mutate(corrected_temp = w_temp_ts-w_temp_gfdl+w_temp_woa)

#plot the temperature time series ranging from January 1961 to December 2010
sb_temp_corr_box |> 
filter(BOX_ID== 2 & layer ==1) |>
ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 1) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue") +
  labs(y = "Sea temperature 0-50 m depth in Box 2") +
  theme_bw()
  
 #calculate differences between datasets
 sb_temp_corr_box <- sb_temp_corr_box |> 
 mutate(diff = corrected_temp-w_temp_ts)
 #print the monthly differences
 head(sb_temp_corr_box, 12)
 
 # Saving datasets
 #write.csv(sb_temp_corr_box, file.path(output,"sb_temp_corr_box.csv"), row.names = FALSE)

```
