---
title: "Creating bias-corrected time series of sea temperature"
author: "Kelly Ortega based on Denisse Fierro- Arcos script"
date: "2024-10-24"
output: html_document
---
In this notebook, we describe the process to create time series of bias corrected temperature following the guidelines from Ortega-Cisneros et al., (2024).

We will use data data downloaded from our FishMIP Input Explorer app, specifically, the data from the 'Model outputs against observations' tab for three case study regions. The data is downloaded as a folder containing four files. You will need to unzip the folder before running this script to access the .parquet files.

#Loading relevant libraries

```{r setup, include=FALSE}
.libPaths("/home/user/R/x86_64-pc-linux-gnu-library/4.3")
library(nanoparquet)
library(dplyr)
library(Rarr)
library(tidyr)
library(terra)
library(sf)
library(tibble)
library(ggplot2)
library(lubridate)
```

### Defining the location to the input files
```{r}
inputpath <- "~/Documents/Fish-MIP/scripts/3a_paper_revised/temp_sbenguela_atlantis/"
```

### Loading parquet files for the Hawai'i based longline fishery downloaded from the FishMIP THREDDS server 
```{r }
sb_g <- file.path(
  inputpath,
  "gfdl-mom6-cobalt2_obsclim_thetao_15arcmin_southern-benguela_mthly_clim_mean_1981_2010.parquet") |> 
  read_parquet()
sb_woa <- file.path(
  inputpath,
  "regridded_woa_southern-benguela_month_clim_mean_temp_1981-2010.parquet") |> 
  read_parquet()
sb_areacello <- file.path(
  inputpath,
  "gfdl-mom6-cobalt2_areacello_15arcmin_southern-benguela_fixed.parquet") |> 
  read_parquet()
sb_depth <- file.path(
  inputpath,
  "gfdl-mom6-cobalt2_obsclim_deptho_15arcmin_southern-benguela_fixed.parquet") |> 
  read_parquet() 
```

### Exploring file contents
```{r }
parquet_column_types(sb_g)
parquet_column_types(sb_woa)
parquet_column_types(sb_areacello)
parquet_column_types(sb_depth)
```

### Wrangling temperature data to calculate climatologies for the whole model area
```{r }
#Plotting temperature data for the upper most layer from GFDL
sb_g |>
  filter(depth == min(depth)) |>
  ggplot(aes(lon, lat, fill = vals)) +
  geom_tile()

#Calculating height of depth bins - This step is needed because the depth value
#includes the depth of the water column up to that grid cell. If weighting by
#depth, we should only include the depth for the grid cell (i.e., height of the
#grid cell).
depth_bins <- sb_g |> 
  filter(depth < 51) |> 
  distinct(depth) |> 
  #Calculating the difference between each depth bin to get the height of the
  #grid cell
  mutate(depth_height = depth-lag(depth, default = 0))


# Calculating GFDL climatologies from 1981-2010
sb_g_clim <- sb_g |>
  #Keeping relevant columns only
  select(lat:vals) |> 
  filter(depth < 51) |>
  rename(temperature = vals) |>
  drop_na(temperature) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |> 
  rename(area = vals) |> 
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  select(!standard_name:units) |> 
  group_by(month) |>
  #Using volume as weights in mean calculation
  summarise(w_temp_gfdl = weighted.mean(temperature, volume, na.rm = TRUE)) |> 
  #Transforming month column to factor
  mutate(month = factor(month, levels = month.name, ordered = T))

#Plotting temperature data for the upper most layer from WOA climatology
sb_woa |>
  filter(depth == min(depth)) |>
  ggplot(aes(lon, lat, fill = vals)) +
  geom_tile()

# Calculating WOA climatologies from 1981-2010 for the 0-50 m depth layer or realm of this model
sb_w_clim <- sb_woa |>
  #Keeping relevant columns only
  select(lat:vals) |> 
  filter(depth < 51) |>
  rename(temperature = vals) |>
  drop_na(temperature) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |> 
  rename(area = vals) |> 
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  select(!standard_name:units) |> 
  #Excludes grid cells that are not present in GFDL
  drop_na(volume) |> 
  group_by(month) |>
  #Using volume as weights in mean calculation
  summarise(w_temp_woa = weighted.mean(temperature, volume, na.rm = TRUE)) |>
  #Transforming month column to factor
  mutate(month = factor(month, levels = month.name, ordered = T))
```

### Loading GFDL obsclim Zarr file downloaded from the FishMIP Input Explorer app for the southern Benguela 
```{r}
sb_temp <- file.path(
  inputpath,
  "gfdl-mom6-cobalt2_obsclim_thetao_15arcmin_southern-benguela_monthly_1961_2010.zarr")
```

### Exploring zarr file contents
```{r }
zarr_overview(sb_temp)

  # Path: thetao
  # Shape: 600 x 35 x 36 x 55
  # Chunk Shape: 600 x 5 x 36 x 55
  # No. of Chunks: 7 (1 x 7 x 1 x 1)
  # Data Type: float32
  # Endianness: little
  # Compressor: blosc
```

### Loading data

We will add the variable name to the end of the `zarr` path we defined above to load all data.
```{r }
temp_data <- read_zarr_array(file.path(sb_temp, "thetao"))
depth <- read_zarr_array(file.path(sb_temp, "depth_bin_m"))
lat <- read_zarr_array(file.path(sb_temp, "lat"))
lon <- read_zarr_array(file.path(sb_temp, "lon"))
time <- read_zarr_array(file.path(sb_temp, "time"))
```

### Checking dimensions of temperature data

```{r }
dim(temp_data)
#File dimensions are 600 (time) 35 (depth) 36 (lat)  55 (lon)

#plot slice 1 for time and depth
image(temp_data[1,1,,])
```

### Creating a data frame from all components

Important to check dimensions of dataset. Note that dimensions need to match the order of dimensions in the temperature data
```{r }
temp_sb <- cbind(expand.grid(time, depth, lat, lon), val = as.vector(temp_data))

#Rename the columns to reflect their contents
names(temp_sb) <- c("time", "depth", "lat", "lon", "temperature")
```

### Calculate monthly temperature time series from GFDL obsclim

```{r}
#Process data for the upper 50 m depth layer 
sb_gfdl <- temp_sb |> 
  filter(depth < 51) |> 
  drop_na(temperature) |>
  mutate(date = as.Date("1961-01-01") %m+% months(round(time/30.417, 0))) |>
  group_by(lat, lon, date) |>
  mutate(w_temp = weighted.mean(temperature, depth, na.rm = TRUE)) |>
  full_join(sb_areacello, by = c("lat", "lon")) |>
  rename(area = vals) |>
  group_by(date) |>
  summarise(wad_temp = weighted.mean(temperature, area, na.rm = TRUE)) |> 
  #Adding month column to perform join with monthly climatological mean data
  mutate(month = factor(month.name[month(date)], levels = month.name, 
                        ordered = T))

#plot the temperature time series ranging from January 1961 to December 2010
ggplot(sb_gfdl, aes(date, wad_temp)) +
  geom_line() +
  labs(y = "Sea temperature 0-50 m depth") +
  theme_bw()
```

### Calculate bias corrected temperature time series 

```{r}
#Process data first depth bin to check we have process data correctly
#Join monthly means with monthly climatologies from GFDL and WOA
sb_temp_corr <- sb_gfdl |> 
  left_join(sb_g_clim, by = "month") |> 
  left_join(sb_w_clim, by = "month") |> 
  #Apply correction
  mutate(corrected_temp = wad_temp-w_temp_gfdl+w_temp_woa)

#plot the temperature time series ranging from January 1961 to December 2010
ggplot(sb_temp_corr, aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 1) +
  geom_line(aes(y = wad_temp), color = "red") +
  labs(y = "Sea temperature 0-50 m depth") +
  theme_bw()
  
 #calculate differences between datasets
 sb_temp_corr <- sb_temp_corr |> 
   mutate(diff = corrected_temp-wad_temp)
 #print the monthly differences
 head(sb_temp_corr, 12)
```

### Calculate climatologies for the model geometry 
```{r }
#Calculating height of each depth bin
depth_bins <- temp_sb |> 
  filter(depth < 501) |> 
  distinct(depth) |> 
  #Calculating the difference between each depth bin to get the height of the
  #grid cell
  mutate(depth_height = depth-lag(depth, default = 0))

# Calculating GFDL time series from 1961-2010 at the lat lon point
sb_g_clims <- temp_sb |>
  filter(depth < 501) |> 
  drop_na(temperature) |>
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  mutate(date = as.Date("1961-01-01") %m+% months(round(time/30.417, 0))) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #may need to convert time into months by dividing by 30.417
  group_by(lon, lat, date, layer) |> 
  summarise(w_temp = weighted.mean(temperature, depth_height, na.rm = TRUE)) |>
  full_join(sb_areacello, by = c("lat","lon")) |>
  select(!standard_name:units) |> 
  rename(area = vals)
```


# Create a mask from shapefile 
**NOTE!!!!!!** - You will need to add the grouping information above, before calculating the weighted means. You should be weighting by volume as in the previous section.

But below you can see how you can get the information about groups

```{r}
#Reading shapefile to perform data extraction for the model polygons
cb <- st_read("~/Documents/abacus/Hydro/model_regions_v3_geo3.shp")

#Create raster from data frame
ras <-  sb_g_clims |>
  ungroup() |>  
  #A single date and layer is needed as a sample
  filter(date == min(date) & layer == min(layer)) |> 
  #Only these three columns are needed to create a raster
  select(lon, lat, area) |> 
  #Create raster
  rast(type = "xyz", crs = "epsg:4326")

#Create raster mask
cb_mask <- rasterize(cb, ras, field = "BOX_ID", background = NA) |> 
  #Transform to data frame
  as.data.frame(xy = T) |> 
  rename(lon = x, lat = y)
#This contains a data frame identifying the group that each cell belongs to

#Join data frames
sb_g_clims |> 
  full_join(cb_mask, by = c("lat", "lon"))
```

