---
title: "Creating bias-corrected time series of sea temperature"
author: "Kelly Ortega-Cisneros, Denisse Fierro-Arcos"
date: "2024-10-24"
output: pdf_document
---
In this notebook, we describe the process to create time series of bias corrected temperature following the guidelines from Ortega-Cisneros et al., (2024).

We will use data data downloaded from our FishMIP Input Explorer app, specifically, the data from the 'Model outputs against observations' tab for three case study regions. The data is downloaded as a folder containing four files. You will need to unzip the folder before running this script to access the `.parquet` files.

# Loading relevant libraries

```{r setup, include=FALSE}
library(nanoparquet)
library(dplyr)
library(Rarr)
library(tidyr)
library(terra)
library(sf)
library(ggplot2)
library(lubridate)
library(cowplot)
```
  
# Loading relevant files
We use the southern Benguela model as an example, but you can replace it with data for any regional model available via the FishMIP Shiny app. You need to ensure that files are downloaded and stored in the `data` folder of this repository prior to running this script.  
  
```{r}
# Temperature data from GFDL
sb_gfdl <- file.path(
  "../data",
  "gfdl-mom6-cobalt2_obsclim_thetao_15arcmin_southern-benguela_mthly_clim_mean_1981_2010.parquet") |> 
  read_parquet()

# Temperature data from WOA
sb_woa <- file.path(
  "../data",
  "regridded_woa_southern-benguela_month_clim_mean_temp_1981-2010.parquet") |> 
  read_parquet()

# Area of grid cells
sb_areacello <- file.path(
  "../data",
  "gfdl-mom6-cobalt2_areacello_15arcmin_southern-benguela_fixed.parquet") |> 
  read_parquet()

# Depth of grid cells
sb_depth <- file.path(
  "../data",
  "gfdl-mom6-cobalt2_obsclim_deptho_15arcmin_southern-benguela_fixed.parquet") |> 
  read_parquet() 

# Regional model shapefile
benguela <- st_read("../data/model_regions_v3_geo3.shp")
```
  
## Exploring file contents
You can use the `nanoparquet` package to explore the contents of the parquet files as shown below. This is useful to understand the structure of the data and the column types. Alternatively, you can use the `str()` function from base R or `glimpse` from `dplyr` to get a summary of the data.  
  
```{r}
parquet_column_types(sb_gfdl)
```

# Visualising temperature data
Plotting temperature data for the upper most layer from GFDL monthly climatology file.
  
```{r }
sb_gfdl |>
  filter(depth == min(depth)) |>
  #Converting months to factor and ordering them prior to plotting
  mutate(month = factor(month, levels = month.name, ordered = T)) |>
  ggplot(aes(lon, lat, fill = vals)) +
  geom_tile()+
  scale_fill_viridis_c(option = "inferno")+
  facet_wrap(~month)

```
  
The plot shows the monthly climatology (1981-2010) of sea surface temperature for the southern Benguela region from the GFDL model. Warmer colours indicate higher temperatures. The warmest months are January to March.  
  
# Processing temperature data
Currently, the depth file includes the depth from the surface to the depth bin, so we will calculate the height of the depth bins. We do this because we need to calculate the volume of the grid cell to use as a weight when calculating the weighted mean for temperature.  

```{r}
depth_bins <- sb_gfdl |> 
  filter(depth < 501) |> 
  distinct(depth) |> 
  #Calculating the difference between each depth bin to get the height of the
  #grid cell
  mutate(depth_height = depth-lag(depth, default = 0))

# Calculating GFDL climatologies from 1981-2010
sb_gfdl_clims <- sb_gfdl |>
  #Keeping relevant columns only
  dplyr::select(lat:vals) |> 
  rename(temperature = vals) |>
  filter(depth < 501) |>
  drop_na(temperature) |>
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |> 
  rename(area = vals) |> 
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  #Keeping relevant columns only
  dplyr::select(!standard_name:units)
  
#Create raster from data frame
ras <-  sb_gfdl_clims |>
  ungroup() |>  
  #A single date and layer is needed as a sample
  filter(layer == min(layer) & month == min(month)) |> 
  #Only these three columns are needed to create a raster
  dplyr::select(lon, lat, area) |> 
  #Create raster
  rast(type = "xyz", crs = "epsg:4326")

#Create raster mask
cb_mask <- rasterize(benguela, ras, field = "BOX_ID", background = NA) |> 
  #Transform to data frame
  as.data.frame(xy = T) |> 
  rename(lon = x, lat = y)
#This contains a data frame identifying the group that each cell belongs to

#Join data frames
sb_gfdl_clims <- sb_gfdl_clims |>
  full_join(cb_mask, by = c("lat", "lon"))

#Finalise the data extraction for the model polygons
sb_gfdl_clim_box <- sb_gfdl_clims |>
  group_by(layer, month, BOX_ID)|>
  #Using volume as weights in mean calculation
  drop_na(temperature) |>
  summarise(w_temp_gfdl = weighted.mean(temperature, volume, na.rm = TRUE)) |> 
  #Transforming month column to factor
  mutate(month = factor(month, levels = month.name)) |>
  arrange(as.numeric(BOX_ID),month)

#Plotting temperature data for the upper most layer from WOA climatology
sb_woa |>
  filter(depth == min(depth)) |>
  ggplot(aes(lon, lat, fill = vals)) +
  geom_tile()

# Calculating WOA climatologies from 1981-2010 for the 0-50 m depth layer or realm of this model
sb_w_clim_box <- sb_woa |>
  #Keeping relevant columns only
  dplyr::select(lat:vals) |> 
  rename(temperature = vals) |>
  filter(depth < 501) |>
  drop_na(temperature) |>
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  #Adding height of grid cell
  left_join(depth_bins, by = "depth") |> 
  group_by(lon, lat, month, layer) |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |>
  rename(area = vals) |> 
  # Joining cb_mask (data frame identifying the box that each cell belongs to) and the WOA dataset
  full_join(cb_mask, by = c("lat", "lon")) |>
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  dplyr::select(!standard_name:units) |> 
  #Excludes grid cells that are not present in GFDL
  drop_na(volume, temperature) |> 
  group_by(layer, month, BOX_ID) |>
  #Using area as weights in mean calculation
  summarise(w_temp_woa = weighted.mean(temperature, volume, na.rm = TRUE)) |>
  #Transforming month column to factor
  mutate(month = factor(month, levels = month.name)) |>
  arrange(BOX_ID, month)
```

### Loading GFDL obsclim Zarr file downloaded from the FishMIP Input Explorer app for the southern Benguela 
```{r}
sb_temp <- file.path(
 "../data",
  "gfdl-mom6-cobalt2_obsclim_thetao_15arcmin_southern-benguela_monthly_1961_2010.zarr")
```
### Exploring zarr file contents
```{r }
zarr_overview(sb_temp)

  # Path: thetao
  # Shape: 600 x 35 x 36 x 55
  # Chunk Shape: 600 x 5 x 36 x 55
  # No. of Chunks: 7 (1 x 7 x 1 x 1)
  # Data Type: float32
  # Endianness: little
  # Compressor: blosc
```

### Loading data

We will add the variable name to the end of the `zarr` path we defined above to load all data.
```{r }
temp_data <- read_zarr_array(file.path(sb_temp, "thetao"))
depth <- read_zarr_array(file.path(sb_temp, "depth_bin_m"))
lat <- read_zarr_array(file.path(sb_temp, "lat"))
lon <- read_zarr_array(file.path(sb_temp, "lon"))
time <- read_zarr_array(file.path(sb_temp, "time"))
```

### Checking dimensions of temperature data

```{r }
dim(temp_data)
#File dimensions are 600 (time) 35 (depth) 36 (lat)  55 (lon)

#plot slice 1 for time and depth
image(temp_data[1,1,,])
```

### Creating a data frame from all components

Important to check dimensions of dataset. Note that dimensions need to match the order of dimensions in the temperature data
```{r }
temp_sb <- cbind(expand.grid(time, depth, lat, lon), val = as.vector(temp_data))

#Rename the columns to reflect their contents
names(temp_sb) <- c("time", "depth", "lat", "lon", "temperature")
```

### Calculate monthly temperature time series from GFDL obsclim

```{r}
#Process data from GFDL obsclim 
sb_gfdl_box <- temp_sb |> 
  filter(depth < 501) |> 
  drop_na(temperature) |>
  mutate(date = as.Date("1961-01-01") %m+% months(round(time/30.417, 0))) |>
  #grouping depth bins into the four depth layers of the southern Benguela Atlantis
  mutate(layer = case_when(depth < 51 ~ 1, depth > 51 & depth < 101 ~ 2, 
                           depth > 100 & depth < 301 ~ 3, depth > 300 ~ 4)) |>
  left_join(depth_bins, by = "depth") |> 
  #Adding area of grid cell
  left_join(sb_areacello, by = c("lat", "lon")) |>
  rename(area = vals) |> 
  # Joining cb_mask (data frame identifying the box that each cell belongs to) and the GFDL time series
  full_join(cb_mask, by = c("lat", "lon")) |>
  #Calculate volume of grid cell
  mutate(volume = depth_height*area) |> 
  dplyr::select(!standard_name:units) |> 
  #Excludes grid cells that are not present in GFDL
  drop_na(volume, temperature) |> 
  group_by(layer, date, BOX_ID) |>
  #Using area as weights in mean calculation
  summarise(w_temp_ts = weighted.mean(temperature, volume, na.rm = TRUE)) |>
  arrange(BOX_ID, date)
 
 
#plot the temperature time series ranging from January 1961 to December 2010
sb_gfdl_box |>
  filter(BOX_ID == 0 & layer == 1)|>
  ggplot(aes(date, w_temp_ts)) +
  geom_line() +
  labs(y = "Sea temperature 0-50 m depth for box 1 and layer 1") +
  theme_bw()
```

### Calculate bias corrected temperature time series 

```{r}
#Join monthly means with monthly climatologies from GFDL and WOA
sb_temp_corr_box <- sb_gfdl_box |> 
  mutate(month = factor(month.name[month(date)])) |> 
  left_join(sb_gfdl_clim_box, by = c("layer", "month", "BOX_ID")) |> 
  left_join(sb_w_clim_box, by = c("layer", "month", "BOX_ID")) |> 
  #Apply correction
  mutate(corrected_temp = w_temp_ts-w_temp_gfdl+w_temp_woa)

#plot the temperature time series ranging from January 1961 to December 2010
sb_temp_corr_box |> 
filter(BOX_ID== 2 & layer ==1) |>
ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 1) +
  geom_line(aes(y = w_temp_ts), color = "red") +
  labs(y = "Sea temperature 0-50 m depth in Box 2") +
  theme_bw()
  
 #calculate differences between datasets
 sb_temp_corr_box <- sb_temp_corr_box |> 
 mutate(diff = corrected_temp-w_temp_ts)
 #print the monthly differences
 head(sb_temp_corr_box, 12)
 
 # Saving datasets
 #write.csv(sb_temp_corr_box, file.path(output,"sb_temp_corr_box.csv"), row.names = FALSE)

```
#Denisse, the code below is the one I use for the figures on our workflow. I dont think they add to the script, so please delete if you think it is best

### Preparing figures for paper

```{r}
# Plotting the shapefile showing the depth layers and boxes
fig4a <- ggplot(data = benguela) +
  geom_sf(aes(fill= as.factor(botz))) +
  labs(fill = "Depth layers", x="", y="") +
  annotate(geom="text", x=23.0, y=-34.2, label= "4", color="black") +
  annotate(geom="text", x=16.52, y=-32.1, label= "11", color="black") +
  guides(fill = guide_legend(reverse=TRUE))+ 
  scale_fill_brewer(palette = "Blues", direction =-1) +
  theme_bw(base_size = 10)+
  theme(panel.grid.minor = element_blank(),panel.grid.major = element_blank(),legend.position = "bottom")

#plot the temperature time series ranging from January 1961 to December 2010 for box 4
box4_l1 <- sb_temp_corr_box |> 
filter(BOX_ID== 4 & layer ==1) |>
  ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 0.8) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue", linewidth =0.8) +
  ggtitle("0-50m depth")+
  labs(x = '', y = 'Sea temperature')  +
  #scale_x_continuous(limits = c(1950, 2010), breaks=seq(1960, 2010, 20), expand = c(0,2)) + 
  scale_y_continuous(limits = c(14,23), breaks=seq(14,23,2), expand = c(0,0.0)) +
  theme_bw(base_size = 14)+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),legend.position = "none", panel.grid.minor = element_blank(),panel.grid.major = element_blank()) 

box4_l2 <- sb_temp_corr_box |> 
  filter(BOX_ID== 4 & layer ==2) |>
  ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 0.8) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue", linewidth =0.8) +
  ggtitle("50-100m depth")+
  labs(x = '', y = '')  +
  scale_y_continuous(limits = c(10,19), breaks=seq(10,19,2), expand = c(0,0.0)) +
  theme_bw(base_size = 14)+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),legend.position = "none", panel.grid.minor = element_blank(),panel.grid.major = element_blank()) 

fig4b<- plot_grid(box4_l1, box4_l2, nrow =1, align = "h")

#plot the temperature time series ranging from January 1961 to December 2010 for box 4
box11_l1 <- sb_temp_corr_box |> 
filter(BOX_ID== 11 & layer ==1) |>
  ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 0.8) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue", linewidth =0.8) +
  ggtitle("0-50m depth")+
  labs(x = '', y = 'Sea temperature')  +
  scale_y_continuous(limits = c(14,21), breaks=seq(14,21,2), expand = c(0,0.0)) +
  theme_bw(base_size = 14)+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),legend.position = "none", panel.grid.minor = element_blank(),panel.grid.major = element_blank()) 

box11_l2 <- sb_temp_corr_box |> 
  filter(BOX_ID== 11 & layer ==2) |>
  ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 0.8) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue", linewidth =0.8) +
  ggtitle("50-100m depth")+
  labs(x = '', y = '')  +
  scale_y_continuous(limits = c(12,17), breaks=seq(12,17,2), expand = c(0,0.0)) +
  theme_bw(base_size = 14)+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),legend.position = "none", panel.grid.minor = element_blank(),panel.grid.major = element_blank()) 

box11_l3 <- sb_temp_corr_box |> 
  filter(BOX_ID== 11 & layer ==3) |>
  ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 0.8) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue", linewidth =0.8) +
  ggtitle("100-300m depth")+
  labs(x = '', y = '')  +
  scale_y_continuous(limits = c(9,13), breaks=seq(9,13,1), expand = c(0,0.0)) +
  theme_bw(base_size = 14)+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),legend.position = "none", panel.grid.minor = element_blank(),panel.grid.major = element_blank()) 

box11_l4 <- sb_temp_corr_box |> 
  filter(BOX_ID== 11 & layer ==4) |>
  ggplot(aes(date)) +
  geom_line(aes(y = corrected_temp), linewidth = 0.8) +
  geom_line(aes(y = w_temp_ts), color = "dodgerblue", linewidth =0.8) +
  ggtitle("300-500m depth")+
  labs(x = '', y = '')  +
  #scale_x_continuous(limits = c(1960, 2010), breaks=seq(1960, 2010, 20), expand = c(0,2)) + 
  scale_y_continuous(limits = c(7,12), breaks=seq(7,12,1), expand = c(0,0.0)) +
  theme_bw(base_size = 14)+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),legend.position = "none", panel.grid.minor = element_blank(),panel.grid.major = element_blank()) 

fig4c <- plot_grid(box11_l1, box11_l2, box11_l3, box11_l4, nrow =1, align = "h", labels = "C")

fig4_upper <- plot_grid(fig4a, fig4b, nrow =1, labels = c("A", "B"), axis ="l")
pdf("~/Documents/Fish-MIP/ISIMIP3a/Data/output/fig4_rev.pdf",width=12,height=8.25)
plot_grid(fig4_upper, fig4c, ncol=1, align = "v", axis = 'l')
dev.off()
```
```

